Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-11-19 manu-draft.qmd
formatting for submission

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```
## Abstract {.unnumbered}

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere. As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$, best estimate $\pm$ 2 standard errors) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$). As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.

@@

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS). Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.

Another limitation of the open-water method is the estimation of the air-water oxygen flux, which is an important term in the oxygen budget. Most approaches estimate this flux using an empirically derived gas transfer velocity that is dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Particularly challenging is that the exchange velocity in estuaries also depends on water depth and tides [@Ho16] and possibly additional factors, such as fetch and the concentration of total suspended solids [@Borges11]. The importance of the air--water gas flux in the oxygen budget opens up the possibility that the oxygen data themselves may be used to determine the gas transfer velocity, an approach that has been applied in streams [e.g., @riley2013].

@@

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air-water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). The equation has dimensions of moles per unit area of time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.

@@

@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange. The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed. $k_w$ also increases with temperature because the film of water at the air-water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). Changes in model implementation were included in the package as additional changes to BASE. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The remaining three parameters $a$ , $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters. The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for BASE, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space. Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004]. Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$. Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14. The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004. The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$. The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. For the prior standad deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14. However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal. Together, the default priors for $a$, $R$, and $b$ are:

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot. Functions in the EBASE R package allow alternative prior distributions from those above.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes. These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida. Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point). Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st [@akaike1969].

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.

@@

@@

## Figures {.unnumbered}

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic data were created using the (a) $a$ parameter to estimate (b) metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($-D$) and the (c) dissolved oxygen time series (including saturation, $C_{sat}$).](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), air-water oxygen flux (D), and light efficiency ($a$) are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: false
title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```

*Running head*: EBASE metabolic model

## Abstract {.unnumbered}

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere. As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean. As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.

@@

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S. Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.

Another limitation of the open-water method is the estimation of the air--water oxygen flux, which is an important term in the oxygen budget. Most approaches estimate this flux using an empirically derived gas transfer velocity that is dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Particularly challenging is that the exchange velocity in estuaries also depends on water depth and tides [@Ho16] and possibly additional factors, such as fetch and the concentration of total suspended solids [@Borges11]. The importance of the air--water gas flux in the oxygen budget opens up the possibility that the oxygen data themselves may be used to determine the gas transfer velocity, an approach that has been applied in streams [e.g., @riley2013].

@@

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). The equation has dimensions of moles per unit area and time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.

@@

@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange. The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed. $k_w$ also increases with temperature because the film of water at the air--water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The parameters $a$, $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters. The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for BASE, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.

Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004]. Although this synthesis focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$. Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.

The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004. The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$. The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. For the prior standard deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14. However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal.

Together, the default prior distributions for $a$, $R$, and $b$ (@fig-priorplot) are:

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Functions in the EBASE R package allow alternative prior distributions from those above.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes. This assessment evaluated if (1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and (2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year (2021) of continuous data from the NERRS reserve in Apalachicola Bay, Florida. Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, 84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, 84.8815$^{\circ}$ W, \~10 km north of Cat Point). Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling [@akaike1969] of the actual data versus time to create a complete dataset from January 1st to December 31st.

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.

@@

::: {#refs}
:::

@@

## Figure Legends {.unnumbered}

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic data were created using the (a) $a$ parameter to estimate (b) metabolic estimates for gross primary production ($P$), ecosystem respiration ($R$), and gas exchange ($D$) and the (c) dissolved oxygen time series (including saturation, $C_{sat}$).](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), air--water oxygen flux (D), and light efficiency ($a$) are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficiency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}
